How to write a PhD in Biological Sciences: a guide for the uninitiated

### 2021-04-15 1.02-useR.Rmd
# Updating sections to chapters

## Reconciling yourself to doing things you've been avoiding {#useR}
Some 30 odd years ago as an undergraduate I was told that I would have to use computers in one of my 2nd year courses. The thought filled me with dread and I remember putting off going into the computer room for as long as I could. It was a steep learning curve, and I struggled. Each time the cursor got to the end of the screen I hit the carriage return button. And each machine had an odd white device attached called a mouse, and I had no idea what that did. All this seems quite preposterous now, but imagine embarking on a PhD in biological sciences without reconciling yourself to using a PC. This is practically where we are now with R [@rcoreteam2021r], and indeed a wealth of other alternative programming platforms.

@@

### And learning lots of other stuff...
The same principle that I’ve written about above can be extended to all sorts of other aspects of your PhD project. You will have to learn new skills, and some of these will be things that you have previously resisted having to do. Allow yourself to be open to all aspects of learning, and all skills. The act of learning opens new pathways in your brain, and using this ‘muscle’ will facilitate future learning exercises.

>>>>

# Reconciling yourself to doing things you've been avoiding {#useR}
Some 30 odd years ago as an undergraduate I was told that I would have to use computers in one of my 2nd year courses. The thought filled me with dread and I remember putting off going into the computer room for as long as I could. It was a steep learning curve, and I struggled. Each time the cursor got to the end of the screen I hit the carriage return button. And each machine had an odd white device attached called a mouse, and I had no idea what that did. All this seems quite preposterous now, but imagine embarking on a PhD in biological sciences without reconciling yourself to using a PC. This is practically where we are now with R [@rcoreteam2021r], and indeed a wealth of other alternative programming platforms.

@@

## And learning lots of other stuff...
The same principle that I’ve written about above can be extended to all sorts of other aspects of your PhD project. You will have to learn new skills, and some of these will be things that you have previously resisted having to do. Allow yourself to be open to all aspects of learning, and all skills. The act of learning opens new pathways in your brain, and using this ‘muscle’ will facilitate future learning exercises.

### 2021-04-15 1.03-lifescientific.Rmd
# Updating sections to chapters

## The scientific project and scientific living {#lifescientific}

@@

### Example: Invasive species 
Alien species are those that have been moved by human agents across biogeographic barriers into parts of the world where they did not evolve. Once these aliens arrive in their new environment, some breed and become established, and after a lag a fraction of these spread and become invasive, often with substantial impacts on the environment and the socio-economic efforts of people in those invaded areas. Invasive species disproportionately affect poorer people in our society, often completely disassociated with the pathways responsible for the introduction and spread. Removing invasive species requires a considerable amount of effort, and unless this is supplied by tax-payers, via governmental policy, falls on those directly impacted. Invasion science, the expanded transdisciplinary area stemming from invasion biology, is required to engage with all stakeholders in invasions, and most importantly to prevent new invasions. Because introductions are made by humans there is a critical need to communicate with those stakeholders along introduction pathways.

@@

### Should society lead the sciences or sciences lead society? 
Of course, the answer is simply that there must be feedback between science and society. As science advances, society learns of these advances and views are advanced. To become incorporated as a societal norm takes time. Over time, views from science that appear radical to society become mainstream and accepted. Think of the views on the earth being round, or more recently the general acceptance that the universe started with the ‘big bang’. The counterculture, also known as post truth, faction in society that espouses views accepted by the mainstream and proved by science beyond reasonable doubt is not a novel phenomenon. Indeed, in the scientific project, we need to accept that these views will persist, but empower the general populace to understand why and how they can be shown to be false. Hence, your view that the earth is not flat is bolstered if you can provide three ways to demonstrate this as fact (pick the ones you like from 7 listed [here](http://crosstalk.cell.com/blog/seven-ways-to-prove-earth-is-round) or [here](https://www.unlv.edu/news/release/round-earth-clues-how-science-proves-our-home-globe)). Otherwise, your assertion that it is not flat will appear as easily denied as a ‘flat-earther’s’ assertion that it is. The power of the scientific argument in society is that it is not simply contradiction or gainsay, but that there is evidence.

@@

### Carl Sagan’s “Baloney detection kit”

@@

### Live your life scientific
>__“Science is more than a body of knowledge, it is a way of thinking”__
>Carl Sagan [-@sagan2011demon]

@@

### Citizen science
For a long time I struggled with the term 'Citizen science' as I reasoned that many such ventures did not involve citizens doing science. To me, most seemed focussed on outsourcing large tasks to willing public participants: i.e. having citizens do some grunt work. Collecting data, or doing a fraction of the analysis is not science. However, in light of the scientific project, I have changed my mind. All of these projects offer citizens a way to participate in scientific projects in a way that was previously unavailable to them. They promote direct communication between scientists and citizens. In this way, they serve to promote the understanding of science to the general public.

>>>>

# The scientific project and scientific living {#lifescientific}

@@

## Example: Invasive species 
Alien species are those that have been moved by human agents across biogeographic barriers into parts of the world where they did not evolve. Once these aliens arrive in their new environment, some breed and become established, and after a lag a fraction of these spread and become invasive, often with substantial impacts on the environment and the socio-economic efforts of people in those invaded areas. Invasive species disproportionately affect poorer people in our society, often completely disassociated with the pathways responsible for the introduction and spread. Removing invasive species requires a considerable amount of effort, and unless this is supplied by tax-payers, via governmental policy, falls on those directly impacted. Invasion science, the expanded transdisciplinary area stemming from invasion biology, is required to engage with all stakeholders in invasions, and most importantly to prevent new invasions. Because introductions are made by humans there is a critical need to communicate with those stakeholders along introduction pathways.

@@

## Should society lead the sciences or sciences lead society? 
Of course, the answer is simply that there must be feedback between science and society. As science advances, society learns of these advances and views are advanced. To become incorporated as a societal norm takes time. Over time, views from science that appear radical to society become mainstream and accepted. Think of the views on the earth being round, or more recently the general acceptance that the universe started with the ‘big bang’. The counterculture, also known as post truth, faction in society that espouses views accepted by the mainstream and proved by science beyond reasonable doubt is not a novel phenomenon. Indeed, in the scientific project, we need to accept that these views will persist, but empower the general populace to understand why and how they can be shown to be false. Hence, your view that the earth is not flat is bolstered if you can provide three ways to demonstrate this as fact (pick the ones you like from 7 listed [here](http://crosstalk.cell.com/blog/seven-ways-to-prove-earth-is-round) or [here](https://www.unlv.edu/news/release/round-earth-clues-how-science-proves-our-home-globe)). Otherwise, your assertion that it is not flat will appear as easily denied as a ‘flat-earther’s’ assertion that it is. The power of the scientific argument in society is that it is not simply contradiction or gainsay, but that there is evidence.

@@

## Carl Sagan’s “Baloney detection kit”

@@

## Live your life scientific
>__“Science is more than a body of knowledge, it is a way of thinking”__
>Carl Sagan [-@sagan2011demon]

@@

## Citizen science
For a long time I struggled with the term 'Citizen science' as I reasoned that many such ventures did not involve citizens doing science. To me, most seemed focussed on outsourcing large tasks to willing public participants: i.e. having citizens do some grunt work. Collecting data, or doing a fraction of the analysis is not science. However, in light of the scientific project, I have changed my mind. All of these projects offer citizens a way to participate in scientific projects in a way that was previously unavailable to them. They promote direct communication between scientists and citizens. In this way, they serve to promote the understanding of science to the general public.
